---
title: "Analysis Report 2: Your Title Here"
author: "Chelsea Noack"
date: "November 10, 2017"
output: github_document
bibliography: references.bib
csl: bioinformatics.csl
---

# Introduction

Smoking is a prevalent habit with dire consequences. Evidence of smoking's effects became evident beginning as early as the 1930's. Since then, it is common knowledge. Not all lung cancers are the same. For example, non small cell lung adenocarcinoma (NSCLC) accounts for 85% of lung cancers. Of those who have NSCLC, only about 15% survive five years post-diagnosis [@ettinger2010non]. Thus, not only does lung cancer clearly have a deadly consequence, but it is the most fatal cancer 

But one component of lung cancer still leaves us worried. While the cancer is preventable by not having a tobacco habit, there is still a genetic component. In other words, even those who have never smoked can still get non small cell lung adenocarcinoma. Of the known carcinogens in cigarettes, they convert to reactive metabolites. If not inactivated, they can bind to cellular DNA and form adducts, a product of added molecules to create a distinct molecular species [@zienolddiny2005polymorphisms]. The large adducts which form typically get repaired, however researchers find that lung cancer patients have low DNA repair capacity. Zienolddiny et al. state that "variant DNA repair genotypes may alter susceptibility to lung cancer", indicating a genetic component within the repair mechanism, making certain persons more susceptible to lung cancer despite smoking or not smoking (2005). 

Thus, if smoking is not only caused by habits, but there is also a genetic component, it leads me to wonder if females and males differ in this genetic component? Can sex be a factor when it comes to the genetic quality of DNA repair mechanisms and thus carry on genes influencing the progression of lung cancer? In addition, will the age of certain females and males be a factor in terms of which stage of lung cancer they get diagnosed? For example, is it primarily younger non-smoker females/males with severe cases of lung cancer, or, more predictably, is it those whom are younger, non-smokers whom have stage 1? 

These questions must take into account the anthropological effects of habits/genetics on females and males. In some countries, smoking is looked down upon more than others. Also in some countries, women and men are expected to have different habits. Thus, in order to answer these questions I will have to look at data from one area. In this case, studies done on Korean patients will be used in order to answer our questions. Korea is known for many prevalent cancers which are not pertained to the lunch, such as stomach, colon/rectal, and thyroid among both sexes. However, lung cancer is still the most deadly, accounting for 15,623 deaths in 2010 [@jung2013cancer]. With this known, I will use data from the Seo et al. and Li et al. studies in order to better answer these questions connecting sex to the genetic components of lung cancer and its severity for non-smokers. 

I hypothesize that those whom are younger non-smokers will have lower cancer stages than those who are older smokers. In addition, I argue that females will have a higher genetic component than men (indicating cancer from non-smoking rather than smoking) given the cultural context. 

# Methods

## Sample origin and sequencing

The data comes from Seo et al.'s 2012 paper *The transcriptional landscape and mutational profile of lung adenocarcinoma*. Seo et al. took fresh surgical samples of normal and cancerous tissues from  200 Korean patients. The patients had primary lung adenocarcinoma and recently underwent lung resection. For their transcriptional study between smokers and non-smokers, which pertains to Li et al.'s study, they used about 70 of those 200 samples. The samples were normal and cancerous tissues from 34 smokers and 34 never-smokers. Within the smoking category was current smoking and those whom have smoked. They also recorded gender, cancer stage, and smoking status from the patients [@seo2012transcriptional]. 

To sequence the samples, they did transcriptome sequencing combine with whole-exome and transcriptome sequencing. Transcriptome sequencing has a past history of detecting somatic mutations in cancer. The combination made their study the first large scale-study of lung adenocarcinoma using RNA sequencing [@seo2012transcriptional]. 

Li et al. (2015) downloaded Seo et al.'s RNAseq data from GEO, Gene Expression Omnibus. In doing so, they could perform a comprehensive pair-wise comparison analysis. From the data, they had 14 billion paired-end sequence reads ranging to 101 bp in length [@li2015rna].  They excluded the data from any patients under the age of 75 for better comparison. Once they confirmed a paired-nature of the RNA-seq data of normal vs. tumor tissue, they used Ensembl GRCh37 Tophat to align the reads to the human genome. To count the reads by genes, they used HTSeq [@li2015rna]. For the differential gene expression analysis, they used R Bioconductor edgeR. Next, Li et al. had to filter the data. Genes with 1 count per million (cpm) in atleast half the sample size were included. From there, we took their edited and filtered data for our analysis. 

## Computational

First, Dr. Napauka Zimmerman downloaded the files from the NCBI Sequence Read archive under the study number ERP001058. Downloaded the extensive amount of data was made easier by the BioMartr package. Biomartr is a package in CRAN which helps retrieve metagenomic data and functional annotation retrieval. It is more polished than NCBI. Next, the files were converted from SRA to fastq. 

Because the files were so massive, Zimmerman had to use Sailfish. Sailfish is a package which quantifies the abundance of previously annotated RNA isoforms from RNA-seq data, and in doing so avoids mapping reads [@patro2014sailfish]. This makes it 20 times faster than alternative methods and still maintains a level of accuracy. This is ideal for processing large amounts of sequencing reads. Once the files were processed, they were put through fastQC, a quality checking program, which eliminated many reads which were very low quality. The QC reports are helpful to see how viable each sequence is for analysis; if the majority of the sequence is above a 30 score, then we consider it to be sequenced well. Typically, the 3' end will deteriorate, which is normal. 

The next step was to put the files through paired-end Trimmomatic. Specifically, it clipped off Illumina adapters and any reads with a quality score less than 20. The trimmed, quality-checked sequences then needed to be aligned. Sailfish efficiently aligned the sequences using k-mers.  As mentioned before, Sailfish is fast; each sample is parsed out. 88 cores were used in an overnight run. Finally, with parsed sequences efficiently aligned, Zimmerman built the tables. Within our table, each row is a human gene name and the column is different samples in all the groups. The table was melted so it included the metadata and original data. Dplyr and ggplot help us analyze the data from the melted tables. 

# Results

In addition to a minimum of 4-5 figures/tables (and associated captions), you should include sufficient text in this section to describe what your findings were. Remember that in the results section you just describe what you found, but you don't interpret it - that happens in the discussion.

```{r load-libraries, message = FALSE, echo = FALSE}
# Be sure to install these packages before running this script
# They can be installed either with the intall.packages() function
# or with the 'Packages' pane in RStudio

# load general-use packages
library("dplyr")
library("tidyr")
library("knitr")
library("ggplot2")
library("magrittr")

# this package allows for the easy inclusion of literature citations in our Rmd
# more info here: https://github.com/crsh/citr
# and here:
# http://rmarkdown.rstudio.com/authoring_bibliographies_and_citations.html
library("citr")
```

```{r load-data, message = FALSE, echo = FALSE}
# load the dataset from a compressed binary file
# it gets loaded as an object called "final_table"
# this has 3.4 million rows...so you will need to be thoughtful about
# how you analyze the data so that you don't overwhelm your laptop
load("output/final_compiled_counts/joined_count_data.RData")

# changes all columns names to lower case so Travis doesn't complain
names(final_table) %<>% tolower

# test that it loaded correctly before proceeding
stopifnot(exists("final_table"))
```

```{r make-meanage-table, echo = FALSE}
# There are many many genes in this dataset, so we can
# subset it down to just a few here to look for interesting patterns
# in the most highly expressed
age_15 <- final_table %>%
  group_by(cancer_stage, gender) %>%
  summarize(mean_age = mean(age_at_diagnosis)) %>%
  arrange(desc(mean_age)) %>%
  head(n = 15)

# then we can use the `kable()` function to make a nicely formatted
# markdown table
age_15 %>%
  kable()
```

**Table 1**: A table of the mean ages for the different categories of cancer stages and gender. The eldest, 75 years old, typically will be a male with stage 4 cancer. The youngest (although not the youngest of all metadata) is 50 years old females with stage 4. 

```{r make-barplot-of-meanage, echo = FALSE}
# this code uses the same data as above, but use it to make a
# barplot - remember geom_col() is just like
# when you use geom_bar(stat = "identity")
age_15 %>%
  ggplot(aes(x = gender,
             y = mean_age,
             fill = cancer_stage)) +
    geom_jitter(width = 0.15, aes(color = cancer_stage))
```

**Figure 1**: A figure showing the abundance of cancer stages against mean ages. We see stage 4 cancer most prevalent in males above 75 years old and in females about 50 years old.

```{r make-smokingstatus-table, echo = FALSE}
# There are many many genes in this dataset, so we can
# subset it down to just a few here to look for interesting patterns
# in the most highly expressed
smoke_30 <- final_table %>%
  group_by(smoking_status, gender, cancer_stage) %>%
  summarize(mean_age = mean(age_at_diagnosis)) %>%
  arrange(desc(mean_age)) %>%
   head(n = 30)

# then we can use the `kable()` function to make a nicely formatted
# markdown table
smoke_30 %>%
  kable()
```

**Table 2**: A more detailed table indicating smoking status of the average oldest and youngest females within the scope of the metadata.

```{r make-barplot-of-smokingstatus, echo = FALSE}
# this code uses the same data as above, but use it to make a
# barplot - remember geom_col() is just like
# when you use geom_bar(stat = "identity")
smoke_30 %>%
  ggplot(aes(x = smoking_status,
             y = mean_age,
             fill = gender)) +
    geom_col(position = "dodge")
```

**Figure 2**: Using a figure to visualize smoking status between females and males based on average ages. 

```{r make-pointplot-of-smokingstatus, echo = FALSE}
# this code uses the same data as above, but use it to make a
# barplot - remember geom_col() is just like
# when you use geom_bar(stat = "identity")
smoke_30 %>%
  ggplot(aes(x = cancer_stage,
             y = mean_age,
             fill = gender)) +
    geom_point(size = 6, aes (shape = gender))
```

**Figure 3**: Another visualization.

```{r make-gene-table, echo = FALSE}
# There are many many genes in this dataset, so we can
# subset it down to just a few here to look for interesting patterns
# in the most highly expressed
top_15 <- final_table %>%
  group_by(gender, genename) %>%
  summarize(mean_count = mean(counts_lengthscaledtpm)) %>%
  arrange(desc(mean_count)) %>%
  head(n = 15)

# then we can use the `kable()` function to make a nicely formatted
# markdown table
top_15 %>%
  kable()
```

**Table 3**: Gene names in top 15.

```{r make-boxplot-of-highly-expressed-genes, echo = FALSE}
# here we just want to pull out the unique gene names and turn
# them into a vector so we can use it below to make a boxplot
# we use the pull() funtion to get this as a vector, just like
# we did when making histograms several weeks ago
top_genes <- top_15 %>%
  ungroup() %>%
  select(genename) %>%
  unique() %>%
  pull()

# now we need to filter from the full data set again, because
# we don't just want summary data, we want all the data in
# order to make boxplots
final_table %>%
  filter(genename %in% top_genes) %>%
  ggplot(aes(x = genename,
             y = age_at_diagnosis,
             fill = gender)) +
    geom_boxplot() +
    facet_wrap(~smoking_status) +
    xlab("Gene Name") +
    ylab("Age of diagnosis per gene") +
    ggtitle("Age of diagnosis per gene by gender and cancer stage") +
    theme_bw() + # simplifies theme
    theme(axis.text.x = # rotates x axis labels vertically
            element_text(angle = 90,
                         hjust = 1))
```

**Figure 4**:

```{r make-barplot-of-highly-expressed-genes, echo = FALSE}
# this code uses the same data as above, but use it to make a
# barplot - remember geom_col() is just like
# when you use geom_bar(stat = "identity")
top_15 %>%
  ggplot(aes(x = gender,
             y = mean_count,
             fill = genename)) +
    geom_col(position = "dodge")
```

**Figure 5**: Genes by gender

# Discussion

**Findings**:  

My analysis of correlation between gender and cancer proclivity, using the Seo et al. and Li et al. data, shows results completely opposite of my hypothesis. 

We see that a high abundance of young female patients had later stage lung cancer while older female patients had early stage lung cancer. Odd. However, similarly to my hypothesis, younger males had early stage lung cancers and older males had late stage lung cancer. Both females and males were similar for stages 1A, 1B and 2A (Figure 1).

First, the conclusion with the males makes sense as it goes along with my hypothesis. I claimed that as you get older, the more likely you are to have later stage lung cancer than younger people. This is based off of uncontrolled cell growth and poor repair mechanisms as you age, a basic consequence of senescence. This is prevalent in this data set for specifically the oldest males having stage 4 lung cancer (Table 1, Figure 1). However, we do not see this pattern all across for males. In fact, we see the average 69 year old male has 2A and 68 year old male has 1B. Where we would expect stage 3B to be, amongst the eldest, is actually in the lowest mean age bracket of 52 years old (Table 1). 

Females faced this same issue, except for stage 4. Thus, it's important to look at smoking status.

Table 2 shows that for the young female Korean patients whom had stage 4, they claim to never smoke. It would be far too simple to argue that it's based on sampling bias. Rather, this is a crucial part of my analysis which will be developed later. Another important facet of Table 2 is that the youngest males with stage 3B cancer also claimed to never smoke. Interesting. While it's not stage 4, it appears that the propensity of never smoking and higher stage of cancer amongst younger patients is clear. Noteably in its accompanying figure, we see an almost equivalent age between females and males for never smokers. The abundance of an "unknown" category is definitely important to note, especially as its make-up is primarily males, however it does not serve a place for my analysis. Figure 3 shows something similar, however shows the cancer stages per mean age. A more distinct cluster of both males and females around the same age are within cancer stages 1B, N/A (differing from the N/A smoking status category), and 2A (Figure 3). We see the most drastic differences in mean age and cancer stage at, again, stage 4 and 3B. 1A also has variance between men and women, just not as significantly. 

Given from the first portion of my analysis, it already clearly debunks my hypothesis that younger females and males will have lower stage cancer, when we see the opposite. This led me to ask which genes are responsible for the trend, which isn't an easy undertaking. 

In Figure 4, we see 

**Reasoning**: 

One reason for this is that most cases are identified in the later stages, as symptoms are more clear (Kanwal et al.). Another possibility is that the genetic component is stronger or just as strong as the enviornmental component. In a twin study done by Ahlbom et al. revealed that smoking and genetics were not completely distinct; that the habit of smoking often affects familial patterns of lung cancer in twins. Twin studies (monozygotic and dyzigotic) have been shown to be previously beneficial when understanding cancer epidemiology [@kanwal2017familial]. 

Lastly, the strange occurence of nonsmoker females, or simply the high abundance of nonsmoker females with lung cancer, indicates another worry; women getting secondhand smoke from their partners. In Korea, Jee et al. found that the rate of lung cancer is non smoking women was higher if their husbands smoked due to its secondhand nature. The results are jarring, however I am not quick to jump to that immediate conclusion. 

# Sources Cited
